# Two key concepts: causality and non-stationarity

**NOTES TO USERS**:   
- notebooks are by nature sequential. Chunks at the end depend on variables defined in the snippets at the beginning: don't forget to proceed in order!   
- only the code is provided. For comments of methods & results, we refer to the book.   
- please report errors!    
- don't forget: models are not optimized and variables are probably not optimally chosen, so *performance is often disappointing*.    

**Step 0**: please make sure the dataset is in your working directory!


The first step is to make sure the required packages are installed.


```{r, message = FALSE, warning = FALSE}
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(lubridate)){install.packages("lubridate")}
if(!require(rpart)){install.packages(c("rpart","rpart.plot"))}
if(!require(randomForest)){install.packages("randomForest")}
if(!require(lmtest)){install.packages("lmtest")}
if(!require(mboost)){install.packages("mboost")}
if(!require(breakDown)){install.packages("breakDown")}
if(!require(CausalImpact)){install.packages("CausalImpact")}
if(!require(xgboost)){install.packages("xgboost")}
```


Next, we activate the relevant packages and load the data. 


```{r, message = FALSE, warning = FALSE}
library(tidyverse)                      # Activate the data science package
library(lubridate)                      # Activate the date management package
library(rpart)                          # Package for the trees 
library(rpart.plot)                     # Package for tree plots
library(randomForest)                   # Package for random forests
library(xgboost)                        # Package for boosted trees
load("data_ml.RData")                   # Load the data
data_ml <- data_ml %>% 
    filter(date > "1999-12-31",         # Keep the date with sufficient data points
           date < "2019-01-01") %>%
    arrange(stock_id, date)             # Order the data
```


We also copy/paste & aggregate some chunks from Chapter 2.


```{r, message = FALSE, warning = FALSE}
stock_ids <- levels(as.factor(data_ml$stock_id)) # A list of all stock_ids
stock_days <- data_ml %>%                        # Compute the number of data points per stock
    group_by(stock_id) %>% summarize(nb = n()) 
stock_ids_short <- stock_ids[which(stock_days$nb == max(stock_days$nb))] # Stocks with full data
returns <- data_ml %>%                           # Compute returns, in matrix format, in 3 steps:
    filter(stock_id %in% stock_ids_short) %>%    # 1. Filtering the data
    dplyr::select(date, stock_id, R1M_Usd) %>%   # 2. Keep returns along with dates & firm names
    spread(key = stock_id, value = R1M_Usd)      # 3. Put in matrix shape 
features <- colnames(data_ml[3:95]) # Keep the feature's column names (hard-coded, beware!)
features_short <- c("Div_Yld", "Eps", "Mkt_Cap_12M_Usd", "Mom_11M_Usd", 
                    "Ocf", "Pb", "Vol1Y_Usd")
data_ml <- data_ml %>% 
    group_by(date) %>%                                   # Group by date
    mutate(R1M_Usd_C = R1M_Usd > median(R1M_Usd),        # Create the categorical labels
           R12M_Usd_C = R1M_Usd > median(R12M_Usd)) %>%
    ungroup() %>%
    mutate_if(is.logical, as.factor)
separation_date <- as.Date("2014-01-15")
training_sample <- filter(data_ml, date < separation_date)
testing_sample <- filter(data_ml, date >= separation_date)
```


We start with Granger causality.  


```{r, message = FALSE, warning = FALSE}
library(lmtest)
x_granger <- training_sample %>%                            # X variable =...
    filter(stock_id ==1) %>%     # ... stock nb 1
    pull(Mkt_Cap_6M_Usd)         # ... & Market cap
y_granger <- training_sample %>%                            # Y variable = ...
    filter(stock_id ==1) %>%     # ... stock nb 1
    pull(R1M_Usd)                # ... & 1M return
fit_granger <- grangertest(x_granger,                       # X variable
                           y_granger,                       # Y variable
                           order = 6,                       # Maximmum lag
                           na.action = na.omit)             # What to do with missing data
fit_granger
```


Then: causal additive models




Below, we build the adjacency matrix pertaining to the small set of predictor variables plus the 1 month ahead return (on the training sample). We use the *CAM* package which has a very simple syntax. The CAM package was removed from CRAN in November 2019 but can still be installed manually. First, download the content of the package: https://cran.r-project.org/web/packages/CAM/index.html. Second, copy it in the directory obtained by typing *.libPaths()* in the console.   
For Rgraphviz and graph, it is advised to follow https://www.bioconductor.org/packages/release/bioc/html/Rgraphviz.html, that is, to use BiocManager.

```{r, message = FALSE, warning = FALSE}
# if(!require(CAM)){install.packages("CAM")} # => does not work anymore!
if(!require(pcalg)){install.packages("pcalg")}
if (!require("BiocManager")){ install.packages("BiocManager")}
BiocManager::install("Rgraphviz")
BiocManager::install("graph")
BiocManager::install("RBGL")
```

Afterwards, the syntax is simple.

```{r, message = FALSE, warning = FALSE, size = "footnotesize", cache = TRUE}
library(CAM)                # Activate the package
data_caus <- training_sample %>% dplyr::select(c("R1M_Usd", features_short))
fit_cam <- CAM(data_caus)   # The main function
fit_cam$Adj                 # Showing the adjacency matrix
```

This gives the adjacency matrix. 

For the sake of completeness, we also provide an implementation of the *pcalg* package.


```{r, message = FALSE, warning = FALSE}
library(pcalg)                                             # Load packages
library(Rgraphviz)
est_caus <- list(C = cor(data_caus),  n = nrow(data_caus)) # Compute correlations
pc.fit <- pc(est_caus, indepTest = gaussCItest,            # Estimate model
             p = ncol(data_caus),alpha = 0.01)
iplotPC(pc.fit)                                            # Plot model
```



Next: structural time-series models

Below, we test if the 100$^{th}$ date point in the sample (April 2008) is a turning point. Arguably, this date belongs to the time span of the subprime financial crisis. We use the *CausalImpact* package which uses the *bsts* library (Bayesian structural time series).

```{r, message = FALSE, warning = FALSE}
library(CausalImpact)
stock1_data <- data_ml %>% filter(stock_id == 1)          # Data of first stock
struct_data <- data.frame(y = stock1_data$R1M_Usd) %>%    # Combine label...
    cbind(stock1_data %>% dplyr::select(features_short))  # ... and features
pre.period <- c(1,100)                                    # Pre-break period (pre-2008)
post.period <- c(101,200)                                 # Post-break period
impact <- CausalImpact(zoo(struct_data), pre.period, post.period)
summary(impact)
#summary(impact, "report")                                # Get the full report (see below)
plot(impact)
```


A detailed report can be obtained with the commented line in the above chunk.



Below: a simple plot of average monthly returns.

```{r, message = FALSE, warning = FALSE}
data_ml %>% 
    mutate(year = year(date)) %>%          # Create a year variable
    group_by(year) %>%                     # Group by year
    summarize(avg_ret = mean(R1M_Usd)) %>% # Compute average return
    ggplot(aes(x = year, y = avg_ret)) + geom_col() + theme_grey()
```


Next, we perform simple regressions (with Mkt_Cap_6M_Usd as the predictor), on a year-by-year basis.


```{r, message = FALSE, warning = FALSE}
data_ml %>%
    mutate(year = year(date)) %>%                           # Create a year variable
    group_by(year) %>%                                      # Group by year
    summarize(beta_cap = lm(R6M_Usd ~ Mkt_Cap_6M_Usd) %>%   # Perform regression
                  coef() %>%                                # Extract coefs
                  t() %>%                                   # Transpose
                  data.frame() %>%                          # Format into df
                  pull(Mkt_Cap_6M_Usd)) %>%                 # Pull coef (remove intercept)
    ggplot(aes(x = year, y = beta_cap)) + geom_col() +      # Plot
    theme_grey()
```







